Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat

### 2023-11-06 manuscript.Rmd
various updates including integration of real chl/sst data

library(tidyverse)
library(pander)
library(fs)
library(here)
library(english)
library(phyloseq) 
library(vegan)
library(ggrepel)
library(here)
library(worrms)
library(BiodiversityR)
```

@@

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau. The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities. Studies examining community structure across environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Specifically, we used ARMS to sample brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011]. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers.

A likely important but under-investigated aspect of coral reef community structure is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. <!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018]. <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018]. Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019]. Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.--> Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (1) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (2) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands? (3) What environmental and/or anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago? (4) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

```{r fig-map, fig.cap='Map of ARMS deployment sites across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue).', results="asis", fig.width=9, fig.height=6, dependson="init"}
library(sf)
library(rnaturalearth)

@@

ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate. Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol. Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).

@@

Potential environmental drivers of shallow crab community structure were assembled from published literature [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]). Due to logistical limitations, environmental factors were only available for shallow samples.

```{r env-table,results="asis", dependson="init"}
print_table(model_terms,"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016",hide="variable")
```

@@

![The four most abundant crab species found across all deployed ARMS units. (A) *Chlorodiella laevissima*. (B) *Perinia tumida*. (C) *Percnon abbreviatum*. (D) *Dynomene hispida*. Photo credit: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}

@@

db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
db_lwr <- dbrda(cc$crab_otus_shallow  ~ 1,data=cc$crab_data_shallow,distance="bray")

@@

# get rid of variables with VIF > 3
db_modl <- reduce_model(db_modl,vif_limit=3)

@@

Environmental variables clustered strongly according to island group (MHI vs NWHI), with the difference driven primarily by human impact, coral cover, and island slope ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]). `r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant and sufficiently uncorrelated. The drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).

```{r fig-rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
termz <- model_terms %>%
  select(variable,display=`Environmental Driver`)

@@

# plot the plotz
rda_plotz + 
  xlab(axis_var[1]) + ylab(axis_var[2])
cat("{#fig:fig_rda}")
```

@@

```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by island group and depth zone. Ellipses represent 95% confidence intervals around sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
# make the sample group look nice
nice <- function(s) {
  str_replace(s,"shallow","Shallow") %>%
    str_replace("main","(MHI)") %>%
    str_replace("nwhi","(NWHI)") %>%
    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
    map_chr(~str_c(.x,collapse=" "))
}

@@

```{r fig-alpha, fig.cap='Box plots of (A,B) species richness and (C,D) Simpson diversity for complete dataset (all sampled depths). (A) Species richness by island group. (B) Species richness by sampling depth. (C) Simpson diversity by island group. (D) Simpson diversity by sampling depth. Notches represent nonparametric confidence intervals about the median.', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
library(patchwork)

@@

```{r fig-alpha2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity by island group for shallow samples only. Notches as in [@fig:fig_alpha].', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
art <- shallow_alpha$richness_table %>% 
  mutate(island_group = str_to_sentence(island_group)) %>%
  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))

@@

pcdata <- cds %>%
  select(where(is.numeric))

@@

ggplot(bp,aes(x=PC1,y=PC2)) + 
  geom_point(aes(fill=island_group),shape=21,color="black",size=4) +
  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color="black",alpha=0.1,show.legend = FALSE, size=0.07) +
  # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
  scale_fill_manual(values=island_group_pal) + 
  geom_segment(data=loadings,aes(x=0,y=0,xend=PC1,yend=PC2),color="black",size=0.5,arrow=arrow(length=unit(0.02,"npc"))) + 
  geom_label(data=loadings,aes(x=PC1+xnudge,y=PC2,label=label,hjust=hjust)) + 
  xlab(str_glue("PC1 [{pc(explained[1])}]")) + 
  ylab(str_glue("PC2 [{pc(explained[2])}]"))
cat("{#fig:fig_pca}")
```

>>>>

library(tidyverse)
library(pander)
library(fs)
library(here)
library(english)
library(phyloseq) 
library(vegan)
library(ggrepel)
library(here)
library(worrms)
library(BiodiversityR)
library(ggtext)
```

@@

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau that make up the Papahānaumokuākea Marine National Monument. The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities. Studies examining community structure across environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Specifically, we used ARMS to sample brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011]. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers.

A likely important but under-investigated aspect of coral reef community structure is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. <!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018]. <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018]. Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019]. Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.--> Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and/or anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

```{r fig-map, fig.cap='Map of ARMS deployment sites across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument (NWHI) is shaded in light blue).', results="asis", fig.width=9, fig.height=6, dependson="init"}
library(sf)
library(rnaturalearth)

@@

ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate. Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol. Specimens were collected under permits [#xxx]{.mark} (NWHI) and [#xxx]{.mark} (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).

@@

Potential environmental drivers of shallow crab community structure were assembled from a study on factors affecting population genetics in Hawai&#x02BB;i, [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]), NASA Physical Oceanograpy Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC). For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental meterial. Due to logistical limitations, environmental factors were collected only for shallow samples ($\le$ 30 m).

```{r env-table,results="asis", dependson="init"}
print_table(model_terms,"env_table","Potential environmental drivers of crab community structure across the Hawaiian Archipelago",hide="variable")
```

@@

![The four most abundant crab species found across all deployed ARMS units. (a) *Chlorodiella laevissima*. (b) *Perinia tumida*. (c) *Percnon abbreviatum*. (d) *Dynomene hispida*. Photo credit: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}

@@

# db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
db_lwr <- dbrda(cc$crab_otus_shallow  ~ 1,data=cc$crab_data_shallow,distance="bray")

@@

# get rid of variables with VIF > 3
# db_modl <- reduce_model(db_modl,vif_limit=8)
db_modl <- reduce_model(db_modl,vif_limit=3)

@@

Environmental variables clustered strongly according to island group (MHI vs NWHI), with the difference driven primarily by human impact, coral cover, and island slope ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant and sufficiently uncorrelated. The drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).

```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands. Ellipses represent 95% confidence intervals around groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
termz <- model_terms %>%
  select(variable,display=`Environmental Driver`)

@@

# plot the plotz
rda_plotz + 
  xlab(axis_var[1]) + ylab(axis_var[2])
cat("{#fig:fig_dbrda}")
```

@@

```{r fig-cap, fig.cap="CAP plot of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
# make the sample group look nice
nice <- function(s) {
  str_replace(s,"shallow","Shallow") %>%
    str_replace("main","(MHI)") %>%
    str_replace("nwhi","(NWHI)") %>%
    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
    map_chr(~str_c(.x,collapse=" "))
}

@@

```{r fig-alpha, fig.cap='Box plots of (a,b) species richness and (c,d) Simpson diversity for complete dataset (all sampled depths). (a) Species richness by island group. (b) Species richness by sampling depth. (c) Simpson diversity by island group. (d) Simpson diversity by sampling depth. Notches represent nonparametric confidence intervals about the median.', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
library(patchwork)

@@

```{r fig-alpha2, fig.cap='Box plots of (a) species richness and (b) Simpson diversity by island group for shallow samples only. Notches as in [@fig:fig_alpha].', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
art <- shallow_alpha$richness_table %>% 
  mutate(island_group = str_to_sentence(island_group)) %>%
  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))

@@

pcdata <- cds %>%
  select(where(is.numeric),-c(lon,lat))

@@

ggplot(bp,aes(x=PC1,y=PC2)) + 
  geom_point(aes(fill=island_group),shape=21,color="black",size=4) +
  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color="black",alpha=0.1,show.legend = FALSE, size=0.07) +
  # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
  scale_fill_manual(values=island_group_pal) + 
  geom_segment(data=loadings,aes(x=0,y=0,xend=PC1,yend=PC2),color="black",size=0.5,arrow=arrow(length=unit(0.02,"npc"))) + 
  geom_label(data=loadings,aes(x=PC1+xnudge,y=PC2,label=label,hjust=hjust)) + 
  xlab(str_glue("PC1 [{pc(explained[1])}]")) + 
  ylab(str_glue("PC2 [{pc(explained[2])}]")) + 
  plotz_theme()
cat("{#fig:fig_pca}")
```
